---
title: "Gov 50 Final Project"
author: "Alex Chou"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## INTRODUCTION

Does screen time affect the mental health of children? The rise of social media apps like Instagram, TikTok, and Snapchat in the 21st century prematurely exposes school-age children to unprecedented content. I hypothesize that the more time children spend on their screens, the worse their mental health. We can also examine this effect in the context of age and gender. For example, girls (especially teenagers) may be more likely to seek mental health treatment as a result of excessive screen time: on their screens, youth are exposed to a variety of triggers, including breaking news about climate change or politics, posts from friends exploring the world, and media produced by content creators who exemplify beauty standards. These conduce feelings of despair, FOMO, or self-hatred. 

Adolescents remain a demographic most susceptible to experiencing mental health disorders like depression and anxiety. According to the CDC, suicide is one of the five leading causes of death among individuals aged 12 - 19. Should screentime be a significant factor that contributes to this phenomena, there is a clear need for society to peel teenagers from their screen. Currently, educational policies embrace technology: many students in high school and middle school are expected to conduct research using internet resources and to turn in assignments online. While this has no doubt made learning more efficient, it is worthwhile to ponder whether there are any alternative strategies that can be widely implemented that encourage students to spend more time away from their screens, practicing hobbies like art, sports, or music.

## DATA SOURCE DISCUSSION


My data set it titled "The Effect of Screentime on the Mental Health of Children." This data is from 2020 and comes from the National Survey of Children's Health. This study measured weekday screen time, defined as time in front of a television, computer, mobile device, or other electronic tablet spent watching programming, playing games, accessing the internet, or using social media. The data in this survey is solicited every year by the U.S. Census Bureau and is funded by the HRSA (Health Resources & Services Administration). Approximately 360,000 addresses were sampled, using an instrument that screens and selects households, and eligible responses were received from just over 67,000. The questionnaire was administered by a professional knowledgeable about the health and health care of the given child. Households were able to complete the NSCH online, on paper, or over the phone in Spanish or English.

Weekday screentime is the independent variable of this study and is measured in individuals ages 17 or younger. The dependent variable is mental health, which is measured by reported instances of mental health treatment or mental health treatment needed. If an adolescent sought or needed mental health treatment, this is considered as (1) in the data set; if the adolescent did not seek and did not need mental health treatment, this was counted as (0) in the data set. Both the independent and dependent variable were measured using survey methods in a cross-sectional study (the survey was distributed at a singular point in time).

```{r}

library(tidyverse)
library(ggplot2)
library(dplyr)

file <- read.csv("Screentime_v_Mental_Health_Dataset.csv")

screen_time <- c("Less than 1 hour", "1 hour", "2 hours", "3 hours", "4 or more hours")

screen_time_v_mental_health <- file |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  ) |>
  select(Screentime, treatment) |>
  group_by(Screentime) |>
  summarize(treatment_mean = mean(treatment, na.rm = TRUE)) |>
  ggplot() +
  geom_col(aes(x = factor(Screentime, screen_time), y = treatment_mean), stat = "identity") +
  labs(x = "Screen Time (hours)",
       y = "Average Need for Mental Health Treatment",
       title = "Screen Time Impact on Mental Health",
       source = "National Survey of Children’s Health (NSCH), 2020")
  
screen_time_v_mental_health

```

The graph clearly shows that adolescents who spend a greater amount of time on their screens are more likely to seek treatment for their mental health. In fact, there is a clear upward trend: as the amount of screen time increases, the likelihood that a given individual has sought treatment increases as well. 

## RESULTS


```{r}

screen_time_gender_v_mental_health <- file |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  ) |>
  select(Screentime, Sex, treatment) |>
  group_by(Screentime, Sex) |>
  summarize(treatment_mean = mean(treatment, na.rm = TRUE))

knitr::kable(screen_time_gender_v_mental_health, digits = 3)
```

When we examine the table that takes into account gender, there are slight deviations between male and female children. For shorter screen times (such as less than one hour, 1 hour, and 2 hours), boys are more likely to seek treatment than their female counterparts that have the same screentime. On the other hand for longer screen times (such as 3 hours and 4 hours or more), girls are more likely to seek treatment than their male counterparts.

```{r}

all_count <- file |>
  group_by(Screentime, Sex) |>
  summarize(n = n()) |>
  ggplot(mapping = aes(x = factor(Screentime, screen_time), y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Number of Respondents",
       title = "Distribution of Respondents by Screentime"
  )
all_count

```

The graph shows that there was significant response from all categories, at least 2000 data points across each division of screen time for each gender (at least 4000 respondents in each independent variable category). 

```{r}

all_regression <- file |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  ) |>
  mutate(
    numeric_screentime = case_when(
      Screentime == "Less than 1 hour" ~ 0.5,
      Screentime == "1 hour" ~ 1,
      Screentime == "2 hours" ~ 2,
      Screentime == "3 hours" ~ 3,
      Screentime == "4 or more hours" ~ 4.5
    )
  )

regression_4 <- lm(treatment ~ numeric_screentime, data = all_regression)

varnames_1 <- c(
  "(Intercept)" = "Intercept",
  "numeric_screentime" = "Hourly Screentime"
)

modelsummary::modelsummary(
    regression_4, 
    statistic = "p.value",
    gof_map = c("nobs", "r.squared", "adj.r.squared"),
    coef_map = varnames_1)

```
Given that the survey asked for categorical responses (e.g. "less than 1 hour" or "4 or more hours"), this variable was turned into numerical times. Therefore, all teenagers in my data analysis were assigned screen times of 0.5 (corresponding to "less than 1 hour"), 1, 2, 3, or 4.5 (corresponding to "4 or more hours") hours. While there may be intrinsic flaws with this conversion (for example, "4 or more hours" includes 8 hours of screen time and "less than 1 hour" includes 15 minutes of screen time), this was a necessary step for synthesizing the data.

We can see from the table above that the intercept is roughly 0.02, which implies that—when all other variables are 0, i.e. when a given individual under 18 years of age consistently does not spend any time on their screen on weekdays, approximately 2 percent of these individuals will still require or seek mental health treatment. The slope is about 0.045, which means that for every increased hour of screen time, approximately 4.5 percent more individuals require or seek mental health treatment. 

The coefficients in this table have very small p-values (both less than 0.001), which is less than the alpha value of 0.05, implying that these results are statistically significant. Since this study was conducted in a cross-sectional format, I do not believe that the relationship between screen time and mental health treatment can be interpreted causally here, though there is certainly a very strong correlation. 

```{r}

teenagers <- file |>
  filter(Age..Years. >= 13) |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  )

teenagers_count <- teenagers |>
  group_by(Screentime, Sex) |>
  summarize(n = n()) |>
  ggplot(mapping = aes(x = factor(Screentime, screen_time), y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Number of Teenage Respondents",
       title = "Distribution of Teenagers (13-17) by Screentime"
  )
teenagers_count

```

The graph above shows that, among teenagers, the vast majority spend at least 2 hours on their screens. In fact, a significant proportion spend 4 or more hours of their weekdays dedicated to screentime. We can also see that, more male respondents than female respondents have screen time that falls in the categories of 2, 3, and 4 + hours, whereas the trend is flipped for the 1 hour or less than 1 hour categories.

```{r}

teenagers_graph <- teenagers |>
  select(Screentime, treatment, Sex) |>
  group_by(Screentime, Sex) |>
  summarize(treatment_mean = mean(treatment, na.rm = TRUE)) |>
  ggplot() +
  geom_col(aes(x = factor(Screentime, screen_time), y = treatment_mean, fill = Sex), position = "dodge", stat = "identity") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Average Need for Mental Health Treatment",
       title = "Screen Time Impact on Mental Health for Teenagers (13-17)",
       source = "National Survey of Children’s Health (NSCH), 2020")
teenagers_graph

```

We can see that there is an upward trend: female teenagers with 4 or more hours of screen time are report twice as high rates of mental health treatment than their counterparts with screen times of less than 1 hour (approximately 0.32 and 0.16, respectively). It is interesting that the rates of mental health treatment decrease for teenagers from screen times of less than 1 hour to screen times of 1 hour, but otherwise there is a clear positive relationship between screen time and average need for mental health for teenagers. 

It is particularly interesting to note that rates of mental health treatment are higher for female teenagers than male teenagers, though the general positive applies for both genders. This is particularly evident for teenagers who spend 2, 3, or 4+ hours on screens. In fact, female teens who spend 4+ hours on screens are 10% more likely to require/seek mentla health treatment than their male teen counterparts.

```{r}

teenagers_regression <- teenagers |>
  mutate(
    numeric_screentime = case_when(
      Screentime == "Less than 1 hour" ~ 0.5,
      Screentime == "1 hour" ~ 1,
      Screentime == "2 hours" ~ 2,
      Screentime == "3 hours" ~ 3,
      Screentime == "4 or more hours" ~ 4.5
    )
  )

regression_1 <- lm(treatment ~ numeric_screentime, data = teenagers_regression)

varnames_1 <- c(
  "(Intercept)" = "Intercept",
  "numeric_screentime" = "Hourly Screentime"
)

modelsummary::modelsummary(
    regression_1, 
    statistic = "p.value",
    gof_map = c("nobs", "r.squared", "adj.r.squared"),
    coef_map = varnames_1)

```

We can see from the table above that the intercept is roughly 0.07, which implies that—when all other variables are 0, i.e. when a given teenager consistently does not spend any time on their screen on weekdays, approximately 7 percent of these teenagers will still require or seek mental health treatment. This is the largest intercept of all three age groups, which suggests that teenagers have the largest baseline need for seeking mental health treatment. The slope is about 0.04, which means that for every increased hour of screen time, approximately 4 percent more teenagers require or seek mental health treatment. This is the greatest slope across the three age groups, which also suggests that teenagers are most susceptible to the effects of screen time on mental health (if there is any causal effect).

The coefficients in this table have very small p-values (both less than 0.001), less than the alpha value of 0.05, which implies that these results are statistically significant. Since this study was conducted in a cross-sectional format, I do not believe that the relationship between screen time and mental health treatment can be interpreted causally here, though there is certainly a very strong correlation.

```{r}

children <- file |>
  filter(Age..Years. <= 6) |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  )

children_count <- children |>
  group_by(Screentime, Sex) |>
  summarize(n = n()) |>
  ggplot(mapping = aes(x = factor(Screentime, screen_time), y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Number of Child Respondents",
       title = "Distribution of Children (0-6) by Screentime"
  )
children_count

```

Unsurprisingly, it appears that children aged 0-6 spend much less time on screens than their teenage counterparts. Here, we can see that the vast majority of children fall into the categories of less than 1 hour, 1 hour, or 2 hours of screen time. 

```{r}

children_graph <- children |>
  select(Screentime, treatment, Sex) |>
  group_by(Screentime, Sex) |>
  summarize(treatment_mean = mean(treatment, na.rm = TRUE)) |>
  ggplot() +
  geom_col(aes(x = factor(Screentime, screen_time), y = treatment_mean, fill = Sex), position = "dodge", stat = "identity") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Average Need for Mental Health Treatment",
       title = "Screen Time Impact on Mental Health for Children (0-6)",
       source = "National Survey of Children’s Health (NSCH), 2020")
children_graph

```

The positive linear relationship between screen time and mental health treatment is quite evident from this graph. It is worth noting that the scale of this graph differs from the one before: children aged 0 - 6 are much less likely to seek or require mental health treatment if at all. It is still clear, however, that children who spend more time on their screens are more likely to seek or require mental health treatment: children who spend 4 or more hours on their screens are 3-4 times more likely to need mental heath treatment compared to their counterparts who spend less than an hour on screens. 

The gender trend found for teens is reversed in children: in general, male children are more likely to need mental health treatment than their female counterparts. For example, among children who spend 2 or 3 hours on their screens, boys are twice as likely to need mental health treatment than their female counterparts (roughly 5% for boys vs roughly 2.5% for girls).

```{r}

children_regression <- children |>
  mutate(
    numeric_screentime = case_when(
      Screentime == "Less than 1 hour" ~ 0.5,
      Screentime == "1 hour" ~ 1,
      Screentime == "2 hours" ~ 2,
      Screentime == "3 hours" ~ 3,
      Screentime == "4 or more hours" ~ 4.5
    )
  )

regression_2 <- lm(treatment ~ numeric_screentime, data = children_regression)

modelsummary::modelsummary(
    regression_2, 
    statistic = "p.value",
    gof_map = c("nobs", "r.squared", "adj.r.squared"),
    coef_map = varnames_1)

```

We can see from the table above that the intercept is barely 0.01, which implies that—when all other variables are 0, i.e. when a given child consistently does not spend any time on their screen on weekdays, only about 1 percent of these children will still require or seek mental health treatment. The slope is about 0.015, which means that for every increased hour of screen time, approximately 1.5 percent more children require or seek mental health treatment. 

The coefficients in this table also have very small p-values (the slope less than 0.001), less than the alpha value of 0.05, which implies that these results are statistically significant. Since this study was conducted in a cross-sectional format, I do not believe that the relationship between screen time and mental health treatment can be interpreted causally here, though there is certainly a very strong correlation.

```{r}

adolescents <- file |>
  filter(Age..Years. > 6 & Age..Years. < 13) |>
  mutate(
    treatment = if_else(Mental.Health.Treatment == "Needs Treatment", 1, 0)
  )

adolescent_count <- adolescents |>
  group_by(Screentime, Sex) |>
  summarize(n = n()) |>
  ggplot(mapping = aes(x = factor(Screentime, screen_time), y = n, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Number of Adolescent Respondents",
       title = "Distribution of Adolescents (7-12) by Screentime"
  )
adolescent_count

```

Overall, most adolescents aged 7 - 12 spend 2 hours, 3 hours, or 4 or more hours on screens. 

```{r}

adolescents_graph <- adolescents |>
  select(Screentime, treatment, Sex) |>
  group_by(Screentime, Sex) |>
  summarize(treatment_mean = mean(treatment, na.rm = TRUE)) |>
  ggplot() +
  geom_col(aes(x = factor(Screentime, screen_time), y = treatment_mean, fill = Sex), position = "dodge", stat = "identity") +
  scale_fill_manual(values = c(Male = "cadetblue", Female = "cyan4")) +
  labs(x = "Screen Time (hours)",
       y = "Average Need for Mental Health Treatment",
       title = "Screen Time Impact on Mental Health for Children (7-12)",
       source = "National Survey of Children’s Health (NSCH), 2020")
adolescents_graph

```

There is a similar upward trend shown here: across both male and female adolescents, an increase in screen time can be correlated with a greater proportion of these adolescents seeking mental health treatment. For individuals in this age category, those who spend 4+ hours on their screens are over 4 times more likely to need mental health treatment (between 20 to 25 percent) compared to those who spend less than 1 hour (about 5 percent).

Trends in the effect of gender on mental health treatment are less clear for this group. Among adolescents who spend 2 or fewer hhours on their screens, males have slightly higher rates of mental health treatment, whereas among adolescents who spend 3 or more hours on their screens, females have higher rates of mental health treatment than their male counterparts.

```{r}

adolescents_regression <- adolescents |>
  mutate(
    numeric_screentime = case_when(
      Screentime == "Less than 1 hour" ~ 0.5,
      Screentime == "1 hour" ~ 1,
      Screentime == "2 hours" ~ 2,
      Screentime == "3 hours" ~ 3,
      Screentime == "4 or more hours" ~ 4.5
    )
  )

regression_3 <- lm(treatment ~ numeric_screentime, data = adolescents_regression)

modelsummary::modelsummary(
    regression_3, 
    statistic = "p.value",
    gof_map = c("nobs", "r.squared", "adj.r.squared"),
    coef_map = varnames_1)

```

We can see from the table above that the intercept is around 0.01, which implies that—when all other variables are 0, i.e. when a given adolescent consistently does not spend any time on their screen on weekdays, only about 1 percent of these children will still require or seek mental health treatment. The slope is 0.023, which means that for every increased hour of screen time, approximately 2.3 percent more adolescents require or seek mental health treatment.

The coefficients in this table also have very small p-values (the p-value of both the intercept and screen time are less than 0.001), less than the alpha value of 0.05, which implies that these results are statistically significant. Since this study was conducted in a cross-sectional format, I do not believe that the relationship between screen time and mental health treatment can be interpreted causally here, though there is certainly a very strong correlation.

## CONCLUSION


The results show a very strong relationship between screen time and mental health of children. There is a clear linear trend depicting that, with more time spent on screens, children are more likely to seek treatment for mental health. There are several limitations to and factors of consideration for this study. For example, this study was conducted during the COVID-19 pandemic and around the time of the 2020 presidential election, both of which may have been perceived as quite serious and emotionally disturbing occurrences, even to children. It is also worth noting that there may have been self-selection bias, given that survey respondents opted into the survey, or response bias, since many parent respondents may be in denial of how much they allow their children to spend online and would consequently under-report this value. With more time and money, it would be interesting to repeat this experiment at a more politically and socially neutral time (in order to minimize the effects of naturally mental health disorder inducing events). Furthermore, we could minimize responder bias by using an app or program that measures screen time objectively (like combining the "screen time" reports on iPhones in settings), rather than relying on parents to estimate off of top of mind.
